---
title: "Genetic Mediation Analysis with genmediation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genetic Mediation Analysis with genmediation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set TRUE when real data are available
)
library(genmediation)
```

## Overview

`genmediation` wraps a lavaan structural equation model that estimates how
**maternal polygenic risk scores (PRS)** for a specific trait mediate the path from
genetic liability to **child trait of interest **, through longitudinal
observed maternal trait.

The analysis follows three steps:

1. **Prepare data** — Z-score standardise PRS variables.  
2. **Build model** — Generate lavaan syntax for interested polygenic risk score.  
3. **Fit & extract** — Run the SEM and pull indirect/total effects.

---

## Step 1 — Load and prepare data

```{r load-data}
library(genmediation)

df <- read.csv("path/to/dataset_genr3_3y.csv")

# Z-score the MDD PRS columns
df <- prepare_prs_data(df, prs_type = "MDD")

# Verify
summary(df[c("prs_qc_MDD", "prs_qc_MDD_m", "prs_qc_MDD_f")])
```

---

## Step 2 — Inspect the model syntax (optional)

```{r model-string}
model_syntax <- build_mdd_model()
cat(model_syntax)
```

You can customise the outcome variable or depression timepoints:

```{r custom-model}
custom_model <- build_mdd_model(
  outcome    = "h_c_extern_1_5y",     # swap outcome
  dep_vars   = c(pre  = "dep_pre_m",
                 t2m  = "dep_2m_m",
                 t6m  = "dep_6m_m",
                 t36m = "dep_36m_m"),
  covariates = c("h_m_age", "h_c_sex", "h_c_age_1_5y")
)
```

---

## Step 3 — Fit the model

```{r fit}
fit <- run_genetic_mediation(
  data      = df,
  prs_type  = "MDD",
  bootstrap = 5000   # reduce to ~200 for quick testing
)
```

---

## Step 4 — Examine results

### Lavaan summary

```{r summary}
print_sem_summary(fit)
```

### Indirect effects only

```{r indirect}
indirect <- extract_indirect_effects(fit)
print(indirect)
```

### Full tidy parameter table

```{r tidy}
all_params <- tidy_sem_results(fit)
head(all_params)

# Export
write.csv(all_params, "sem_mdd_results.csv", row.names = FALSE)
```

---

## PPD model

Simply switch `prs_type`:

```{r ppd}
df_ppd <- prepare_prs_data(df, prs_type = "PPD")

fit_ppd <- run_genetic_mediation(
  data     = df_ppd,
  prs_type = "PPD",
  bootstrap = 5000
)

extract_indirect_effects(fit_ppd)
```

---

## Session info

```{r session}
sessionInfo()
```
